\section{Possible Attacks}

Since this scheme of quantum money from knots is based on conjectures, this
points the way to several future attacks.

\begin{enumerate}
\item Reverse engineering from the Alexander polynomial $p$. If the attacker
could easily find all equivalent grid diagrams to a single "canonical" grid
diagram corresponding to the serial number, he could create the correctly
weighted superposition to duplicate $\ket{\$_p}$. Call
$\mathcal{G}_{2\overline{D}}$ the set of all grid diagrams of up to size
$2\overline{D}$.
The attacker could enumerate over
exponentially many grid moves within $\mathcal{G}_2\overline{D}$,
or he could enumerate over all grids in $\mathcal{G}_2\overline{D}$ assuming
he has an oracle for determining if two grid diagrams are equivalent.
Both would take exponential time.
Alternatively, suppose that a polynomial time algorithm exists to extract from
each grid diagram
(encoded in $O(\overline{D}^2)$
bits) either the moves to get to all but exponentially many equivalent
diagrams or (even more unlikely) the equivalent diagrams themselves.
We would have to be able to upper bound the number of equivalent grid diagrams to
also be $O(\overline{D}^2)$, which we currently don't know how to do.
\item
It turns out that a certain class of states can be efficiently copied, which
includes the eigenstates of some classical reversible circuit.
A special case of this
are the states $\ket{\psi_{n,k}}$ used to enact arbitrary controlled phase
rotations in the quantum compiling algorithm of Kitaev, Shen, and Vyalyi
\cite{KSV02}, where the state to copy (say $\ket{\psi_{n,1}}$ which is hard
to produce) is of the
same form as the "empty" register to hold the new copy ($\ket{\psi_{n,0}}$, which
is easy to produce). As in
\begin{displaymath}
\ket{\psi_{n,1}}\otimes\ket{\psi_{n,0}} \rightarrow
\ket{\psi_{n,1}}\otimes\ket{\psi_{n,1}}
\end{displaymath}
If it turns out that weighted superpositions of equivalent
grid diagrams are within that class of states, then the valid money state
$\ket{\$_p}$ could be copied into an entangled pair
$\ket{\$_p} \otimes \ket{\$_p}$
can somehow be used toese are somehow encoded in the grid diagram itself,
waiting to be extracted.
he might be able to somehow extractEven if he is given an
oracle to 
come up with a "canonical" grid diagram corresponding to the serial number and

\end{enumerate}